The evolution of gene expression in the term placenta of viviparous mammals

### 2015-10-14 manuscript/manuscript.Rnw
add the tables of top genes for all of the organisms we have placenta data for

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Common Name        & Reference & Read Length  \\
    \midrule
    \textit{Ateles fusciceps}     & Spider monkey      & \textit{Homo sapiens} & 100 \\
    \textit{Bos taurus}           & Cow                & UMD3.1     & 76 \\
    \textit{Canis familiaris}     & Dog                & CanFam3.1  & 151 \\
    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & 101 \\
    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & 76 \\
    \textit{Homo sapiens}         & Human              & GRCh38     & 76, 151 \\
    \textit{Monodelphis domestica} & Opossum           & BROAD05    & 76 \\
    \textit{Mus musculus}         & Mouse              & GRCm38     & 36,44,151 \\
    \textit{Nannospalax galili}   & Blind mole rat     & \textit{Mus musculus} & 100 \\
    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & 151 \\
    \textit{Spalax carmeli}       & Blind mole rat     & \textit{Mus musculus} & 100 \\
    \bottomrule
  \end{tabular}
  \caption{Reference genomes for aligned species}
  \label{tab:ref_genome_aligned_species}
\end{table}

@@

<<top.genes.function,echo=FALSE,results="hide">>=
top.genes.table <- function(species.name,n=50) {
    temp <- 
        combined.fpkm[species==species.name &
                           !(!is.na(human_name) & 
                                 human_name %in% housekeeping.genes.superset[,gene_short_name]) &
                               !grepl("((RNA\\d|MT)-|^RPS\\d+$)",human_name) &
                                   !grepl("(^MT-|^RP[LS]\\d+[ABCD]?(?:-.+)?$|rRNA$)",gene_short_name)
                      ][order(-mean_fpkm)][,list(name_or_id,
                                                 human_name,mean_fpkm)]
    temp <- temp[!duplicated(name_or_id)][1:n]
    setnames(temp,c("Gene name","Human Name","FPKM"))
    if (species.name=="homo sapiens") {
        temp <- temp[,-2,with=FALSE]
    }
    return(temp)
}
gene.expression <- function(species.name,gene,complete=FALSE) {
    fpkm <- combined.fpkm[species==species.name &
                              gene_short_name==gene,
                          mean_fpkm]

@@

\begin{wraptable}{R}{0.3\textwidth}
  \tiny
<<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
top.human.genes <- top.genes.table(species.name="homo sapiens")
print(xtable(top.human.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.human.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Homo sapiens}}
\label{tab:top50_human_placenta}
\end{wraptable}

@@

\begin{wraptable}{L}{0.3\textwidth}
  \tiny
<<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
top.bonobo.genes <- top.genes.table(species.name="pan paniscus")
print(xtable(top.bonobo.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.bonobo.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Pan paniscus}}
\label{tab:top50_bonobo_placenta}
\end{wraptable}

@@

\begin{wraptable}{R}{0.3\textwidth}
  \tiny
<<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
print(xtable(top.spider.monkey.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.spider.monkey.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Ateles fusciceps}}
\label{tab:top50_spider_monkey_placenta}
\end{wraptable}

@@

\begin{wraptable}{R}{0.3\textwidth}
  \tiny
<<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
top.dog.genes <- top.genes.table(species.name="canis familiaris")
print(xtable(top.dog.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.dog.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Canis familiaris}}
\label{tab:top50_dog_placenta}
\end{wraptable}

@@

\begin{wraptable}{L}{0.3\textwidth}
  \tiny
<<top.50.cow.expressed.genes,echo=FALSE,results="asis">>=
top.cow.genes <- top.genes.table(species.name="bos taurus")
print(xtable(top.cow.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.cow.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Bos taurus}}
\label{tab:top50_cow_placenta}
\end{wraptable}

@@

\begin{wraptable}{L}{0.3\textwidth}
  \tiny
<<top.50.elephant.expressed.genes,echo=FALSE,results="asis">>=
top.elephant.genes <- top.genes.table(species.name="loxodonta africana")
print(xtable(top.elephant.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.elephant.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Loxodonta africana}}
\label{tab:top50_elphant_placenta}
\end{wraptable}

@@

\begin{wraptable}{L}{0.3\textwidth}
  \tiny
<<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
top.opossum.genes <- top.genes.table(species.name="monodelphis domestica")
print(xtable(top.opossum.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.opossum.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Monodelphis domestica}}
\label{tab:top50_opossum_placenta}
\end{wraptable}

@@

@@

Annexins?

TIMP3 -- maybe in pre-eclampsia (23172037)
SERPINH1? with TIMP3 
% http://www.proteinatlas.org/ENSG00000149257-SERPINH1/tissue/placenta

Why are RPL17 and RPS26 here? They are ribosomal proteins. These
should be missing now.

>>>>

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Common Name        & Reference & Read Length  \\
    \midrule
    \textit{Ateles fusciceps}     & Spider monkey      & \textit{Homo sapiens} & 100 \\
    \textit{Bos taurus}           & Cow                & UMD3.1     & 75 \\
    \textit{Canis familiaris}     & Dog                & CanFam3.1  & 150 \\
    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & 100 \\
    \textit{Equus caballus}       & Horse              & EquCab2    & 100 \\ 
    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & 75 \\
    \textit{Homo sapiens}         & Human              & GRCh38     & 75, 150 \\
    \textit{Monodelphis domestica} & Opossum           & BROAD05    & 75 \\
    \textit{Mus musculus}         & Mouse              & GRCm38     & 35,44,150 \\
    \textit{Nannospalax galili}   & Blind mole rat     & \textit{Mus musculus} & 100 \\
    \textit{Ovis aries}           & Sheep              & Oar_v3.1   & 151 \\
    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & 150 \\
    \textit{Spalax carmeli}       & Blind mole rat     & \textit{Mus musculus} & 100 \\
    \textit{Sus scrofa}           & Wild Boar          & Sscrofa10.2 & 51 \\
    \bottomrule
  \end{tabular}
  \caption{Reference genomes for aligned species}
  \label{tab:ref_genome_aligned_species}
\end{table}

@@

<<top.genes.function,echo=FALSE,results="hide">>=
top.genes.table <- function(species.name,n=50) {
    temp <- 
        combined.fpkm[species==species.name &
                           !(!is.na(human_name) & 
                                 human_name %in% housekeeping.genes.superset[,gene_short_name]) &
                               !grepl("((RNA\\d|MT)-|^RP[SL]\\d+$)",human_name) &
                                   !grepl("(^[Mm][Tt]-|^R[pP][lsLS]\\d+[abcdABCD]?[l]?(?:-.+)?$|rRNA$)",gene_short_name)
                      ][order(-mean_fpkm)][,list(name_or_id,
                                                 human_name,mean_fpkm)]
    temp <- temp[!duplicated(name_or_id)][1:n]
    ## remove any genes which have a null name_or_id; they are probably from unassembled contigs
    temp <- temp[!(is.na(name_or_id)|name_or_id==""),]
    setnames(temp,c("Gene name","Human Name","FPKM"))
    if (species.name=="homo sapiens") {
        temp <- temp[,-2,with=FALSE]
    }
    return(temp)
}
gene.expression <- function(species.name,gene,complete=FALSE) {
    fpkm <- combined.fpkm[species==species.name &
                              gene_short_name==gene,
                          mean_fpkm]

@@

\begin{wraptable}{R}{0.3\textwidth}
  \tiny
<<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
top.human.genes <- top.genes.table(species.name="homo sapiens")
print(xtable(top.human.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.human.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not
  ribosomal, mitochondrial, or a housekeeping gene in \textit{Homo
    sapiens}}
\label{tab:top50_human_placenta}
\end{wraptable}

@@

\begin{wraptable}{L}{0.3\textwidth}
  \tiny
<<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
top.bonobo.genes <- top.genes.table(species.name="pan paniscus")
print(xtable(top.bonobo.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.bonobo.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Pan paniscus}}
\label{tab:top50_bonobo_placenta}
\end{wraptable}

@@

\begin{wraptable}{R}{0.3\textwidth}
  \tiny
<<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
print(xtable(top.spider.monkey.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.spider.monkey.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Ateles fusciceps}}
\label{tab:top50_spider_monkey_placenta}
\end{wraptable}

@@

\begin{wraptable}{R}{0.3\textwidth}
  \tiny
<<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
top.dog.genes <- top.genes.table(species.name="canis familiaris")
print(xtable(top.dog.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.dog.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Canis familiaris}}
\label{tab:top50_dog_placenta}
\end{wraptable}

@@

\begin{wraptable}{L}{0.3\textwidth}
  \tiny
<<top.50.cow.expressed.genes,echo=FALSE,results="asis">>=
top.cow.genes <- top.genes.table(species.name="bos taurus")
print(xtable(top.cow.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.cow.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Bos taurus}}
\label{tab:top50_cow_placenta}
\end{wraptable}

@@

\begin{wraptable}{L}{0.3\textwidth}
  \tiny
<<top.50.elephant.expressed.genes,echo=FALSE,results="asis">>=
top.elephant.genes <- top.genes.table(species.name="loxodonta africana")
print(xtable(top.elephant.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.elephant.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Loxodonta africana}}
\label{tab:top50_elphant_placenta}
\end{wraptable}

@@

\begin{wraptable}{L}{0.3\textwidth}
  \tiny
<<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
top.opossum.genes <- top.genes.table(species.name="monodelphis domestica")
print(xtable(top.opossum.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.opossum.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Monodelphis domestica}}
\label{tab:top50_opossum_placenta}
\end{wraptable}

@@

\subsection{Mouse Placenta Transcriptome}

\begin{table}[h]
  \tiny
<<top.50.mouse.expressed.genes,echo=FALSE,results="asis">>=
top.mouse.genes <- top.genes.table(species.name="mus musculus")
print(xtable(top.mouse.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.mouse.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Mus musculus}}
\label{tab:top50_mouse_placenta}
\end{table}

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \textit{Mus musculus} are shown in
\fref{tab:top50_mouse_placenta} with a list of all highly expressed
genes in Supplemental \fref{tab:all_mouse_placenta}.

% No idea what mir-6236 is doing

% AC099934.1 probably overlaps with Rn7sl (part of SRP?)

% AC157822.1 overlaps with Rn7s2

% AC122546.1 = no idea

% Gm3788 = psuedogene (Thymosin beta 10)

\DLA{Why are there so many psuedogenes in the mouse? Maybe because of
  the short read length?}

\subsection{\textit{Nannospalax galili} (Blind Mole Rat) Placenta Transcriptome}

\begin{table}[h]
  \tiny
<<top.50.n.galili.expressed.genes,echo=FALSE,results="asis">>=
top.galili.genes <- top.genes.table(species.name="nannospalax galili")
print(xtable(top.galili.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.galili.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Nannospalax galili}}
\label{tab:top50_n_galili_placenta}
\end{table}

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \textit{Nannospalax galili} are shown in
\fref{tab:top50_n_galili_placenta} with a list of all highly expressed
genes in Supplemental \fref{tab:all_n_galili_placenta}.

\subsection{\textit{Spalax carmeli} (Blind Mole Rat) Placenta Transcriptome}

\begin{table}[h]
  \tiny
<<top.50.s.carmeli.expressed.genes,echo=FALSE,results="asis">>=
top.carmeli.genes <- top.genes.table(species.name="spalax carmeli")
print(xtable(top.carmeli.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.carmeli.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Spalax carmeli}}
\label{tab:top50_s_carmeli_placenta}
\end{table}

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \textit{Spalax carmeli} are shown in
\fref{tab:top50_s_carmeli_placenta} with a list of all highly expressed
genes in Supplemental \fref{tab:all_s_carmeli_placenta}.

\subsection{Wild Boar Placenta Transcriptome}

\begin{table}[h]
  \tiny
<<top.50.pig.expressed.genes,echo=FALSE,results="asis">>=
top.pig.genes <- top.genes.table(species.name="sus scrofa")
print(xtable(top.pig.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.pig.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Sus scrofa}}
\label{tab:top50_pig_placenta}
\end{table}

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \textit{Sus scrofa} are shown in
\fref{tab:top50_pig_placenta} with a list of all highly expressed
genes in Supplemental \fref{tab:all_pig_placenta}.

\subsection{Horse Placenta Transcriptome}

\begin{table}[h]
  \tiny
<<top.50.horse.expressed.genes,echo=FALSE,results="asis">>=
top.horse.genes <- top.genes.table(species.name="equus caballus")
print(xtable(top.horse.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.horse.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Equus caballus}}
\label{tab:top50_horse_placenta}
\end{table}

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \textit{Equus Caballus} are shown in
\fref{tab:top50_horse_placenta} with a list of all highly expressed
genes in Supplemental \fref{tab:all_horse_placenta}.

\subsection{Sheep Placenta Transcriptome}

\begin{table}[h]
  \tiny
<<top.50.sheep.expressed.genes,echo=FALSE,results="asis">>=
top.sheep.genes <- top.genes.table(species.name="ovis aries")
print(xtable(top.sheep.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.sheep.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Ovis aries}}
\label{tab:top50_sheep_placenta}
\end{table}

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \textit{Ovis aries} are shown in
\fref{tab:top50_sheep_placenta} with a list of all highly expressed
genes in Supplemental \fref{tab:all_sheep_placenta}.

@@

SERPINH1? with TIMP3 
% http://www.proteinatlas.org/ENSG00000149257-SERPINH1/tissue/placenta
